Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells

### 2024-07-04 Manuscript.tex
corrections 1

\begin{abstract}
This study provides a reassessment of the use of Magnesium to Calcium
(Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface
temperatures (SST). While \emph{Patella depressa} along the Spanish
shoreline has repeatedly produced reliable correlations between SST and
Mg/Ca ratios, this is not the case for other patelloid species.
\emph{Patella vulgata}, \emph{Nacella deaureata}, and \emph{Nacella
magellanica} have been studied using Mg/Ca with mixed or contrary
results. In this study, we present elemental maps of various such
species together with stable oxygen isotope values that confirm a good
relationship with SST. Our dataset also reassesses a specimen which was
previously unsuccessful in providing significant correlations between
\(\delta\)\textsuperscript{18}O and Mg/Ca ratios. By reassessing these
species and including modern and archaeological samples (n=12) from
three patelloid species (\emph{P. vulgata}, \emph{N. deaureata}, and
\emph{N. magellanica}) we further add to the growing set of evidence for
the reliable use of Mg/Ca ratios to detect palaeotemperature change. As
a result, these species can in the future serve as a means to determine
ontogenetic age and season of capture as well as to reveal locations of
interest within the growth record (i.e.~annual temperature minima and
maxima) for the targeted analysis using \(\delta\)\textsuperscript{18}O
or clumped oxygen isotope analysis.
\end{abstract}

@@

Limpet shells are commonly found within archaeological sites and and
past shorelines due to their robust carbonate structure and the
long-term use of limpets as a marine food source \citep{Robson2023-oz}.
They have successfully been studied in the past in the context of
coastal subsistence economies, site occupation on seasonal
\citep{Shackleton1973-ij, Parker2018-wf, Bosch2018-ud} or long-term
scale \citep{Ortiz2015-mr}, as well as palaeotemperature
\citep{Fenger2007-gf, Surge2012-ba, Wang2012-ee, Colonese2012-ct, Ferguson2011-zl}.
Determining past sea surface temperature (SST) change largely relies on
the measurement of \(\delta\)\textsuperscript{18}O-values within the
calcitic parts of the shell, but attempts have been made to also use
elemental ratios, such as magnesium to calcium (Mg/Ca)
\citep{Graniero2017-io, Ferguson2011-zl}, to have an alternative
measure, that potentially provides a more accurate SST estimate than
\(\delta\)\textsuperscript{18}O-values, which are also affected by
changes in salinity. In addition, the data acquisition of elemental
ratios --- either through laser-ablation-isotope-ratio-mass-spectrometry
(LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much
faster and cost-effective, increasing the number of specimens that can
be studied overall \citep{Durham2017-fh, Hausmann2023-ih}.

Problematically, links between Mg/Ca ratios and SST changes in many
marine mollusc shells, including limpets, have shown to be unreliable
\citep{Surge2008-ri, Wanamaker2008-zl, Schone2010-yl, Freitas2012-tx, Graniero2017-io, Poulain2015-dg, Vihtakari2017-wd}.
This is particularly the case where there is little available additional
information on metabolic processes, organic components of the shell
matrix, intra-increment and intra-shell variability, and growth rates,
which can independently and unpredictably affect Mg/Ca ratios and
confound their interpretation as temperature proxy. Confusingly, in some
cases multiple different temperature equations have been found for the
same species (see Freitas \emph{et al}. \citeyearpar{Freitas2012-tx} and
Vihtakari \emph{et al}. \citeyearpar{Vihtakari2017-wd} and references
therein). Coeval specimens sharing one locality can also show
differences in their relation to SST \citep{Hausmann2019-fi}. Where the
use of Mg/Ca as a palaeotemperature proxy was successful, anomalous
patterns in some specimens had still to be filtered out by hand,
reducing the overall robustness of the results of those successful
studies \citep{Ferguson2011-zl}.

While recent research particularly of \emph{Patella} sp. in the
Mediterranean and Southwest Europe have provided promising results
\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
There remains a lack of clarity for Atlantic limpet species,
particularly since it was last shown here, that they are not reliable
recorders of palaeotemperature \citep{Graniero2017-io}.

In this study we will repeat and expand the analysis of Atlantic limpets
to determine the reliability of Mg/Ca ratios as palaeotemperature
proxies. To do this, we sampled a set of previously published and
unpublished limpet specimens dating to modern and archaeological
contexts using LIBS. LIBS allows us to carry out 2D imaging of entire
shell sections, which helps us to navigate the complex elemental
structure of the shell and better separate the external,
temperature-related changes from the internal and less understood
factors that influence the Mg/Ca ratio. By working with already
published datasets, we were able to simultaneously avoid costs for new
high-resolution \(\delta\)\textsuperscript{18}O-data, to use real-world
examples, and to also provide pilot-data for areas of existing research
interest. Generally establishing the usefulness and reliability of Mg/Ca
as SST proxy in limpets should help to provide a platform for such
future research and an important stepping stone to a better
understanding of elemental ratios in other marine mollusc shells.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2449}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1429}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1633}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2041}}@{}}
\caption{Overview of the modern and archaeological limpet specimens
analysed in this study.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Cambaceres Bay
& ND-1016-3 & Stable oxygen \\
& & & - Tierra del & ND-1016-4 & and carbon \\
& & \emph{N. magellanica} & Fuego (AR) & NM-1016-1 & isotope analysis \\
& & & & NM-1016-3 & \\
& & & & & \\
& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - &
ORK-LT5 & Stable oxygen \\
& & & Westray (UK) & & and carbon \\
& & & & & isotope analysis; \\
& & & & & Mg, Li, Sr, Ca. \\
& & & & & \\
Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
Bay - & QG1-7188-1 & Stable oxygen \\
& & & Westray (UK) & QG1-7189-2 & and carbon \\
& & & & QG2-1061-1 & isotope analysis \\
& & & & QG2-1064-1 & \\
& & & & QG2-7180-1 & \\
& & & & QG2-7180-2 & \\
& & & & QG1-7246-1 & \\
\end{longtable}

@@

There are two main locations that we sourced specimens from: the Beagle
Channel in Tierra del Fuego (Argentina) and the island of West Ray in
Orkney (UK). Four modern specimens were collected from the Beagle
Channel in Cambaceres Bay in October 2016. The area receives around 570
mm precipitation but the channel waters are predominantly influenced by
marine currents and in the period of October 2015 to October 2016 stayed
around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
August to 10ºC in February (Figure~\ref{fig-SSTs} - red).

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
August 2009 and has previously been analysed using stable carbon and
oxygen isotopes, as well as LA-ICP-MS \citep{Graniero2017-io}. Westray
lies about 70 km north of the Scottish mainland and experiences
virtually no freshwater input with salinity values of 34.0 -- 34.5 psu
\citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in March to
13.8ºC in August (Figure~\ref{fig-SSTs} - blue).

@@

We also studied 7 archaeological specimens of \emph{P. vulgata} coming
from Quoygrew on Westray dating to 900 -- 1200 CE. These shells have
previously been studied to characterise seasonal temperature change
during the Medieval Climate Anomaly and the Little Ice Age
\citep{Surge2012-ba}. Together they show a range of lifespans, with some
comparatively long for their size (QG1-7246-1, 12 years within 16 mm of
growth) making them particularly susceptible to time-averaging. That
said, the majority are under 5 years old.

@@

Mg/Ca elemental imaging was carried out using Laser Induced Breakdown
Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany)
using previously published methods \citep{Hausmann2023-ih}. These
involve the ablation of the shell section using an infrared (1064 nm)
laser (1--2 mJ, 100 Hz) onto a surface area of 20--30 µm to generate a
plasma plume. This plume emits light, which is measured using a
synchronised spectrometer (delay time = 0.5 µs, integration time = 1
µs). The resulting light spectrum quantifies the emission lines of
Magnesium (MgII; 279.553 nm) and Calcium (CaII; 315.887 nm) to determine
their intensity ratio. While these two peaks alone do not represent the
molar concentrations of both elements (as opposed to calibration free
LIBS; e.g. \citep{Martinez-Minchero2022-jz}), the intensity ratio is
linearly correlated to the molar concentration \citep{Hausmann2017-oa}
and can be used as a reliable indicator of Mg/Ca changes within the
shell carbonate. All values reported here at in arbitrary units based on
the ratio of intensities of the peaks above.

@@

\subsubsection{Oxygen isotopes}\label{oxygen-isotopes}

High-quality oxygen isotope values (\(\delta\)\textsuperscript{18}O)
were acquired from previous publications on \emph{P. vulgata}
\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
the stable isotope analysis methodology can be found. In short, all
shells were sectioned along the main growth axis to expose the internal
growth structure. From the exposed section carbonate powder samples were
acquired using a micromill with resolutions of 100--200 µm and by
targeting the calcitic M+2 layer of the shells. Analytical precision of
the isotope values was consistently at 0.05--0.10‰ and values are
reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee
Belemnite) standard.

@@

The elemental imaging of \emph{P. vulgata} shells consistently showed
repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca
intensity ratios indicating and low ratios indicating low temperature
periods. As examples, Figure~\ref{fig-Pat_LIBS} shows the
2D-distribution of elemental ratios across the entire section of ORK-LT5
and the preserved anterior side of QG1-7188-1 (additional maps can be
found in the supplementaries). The repeating patterns are found across
the calcitic layers that are exterior to the myostracum (layers m+2 and
m+3) as well as the layers that are interior (m-2)
\citep{Fenger2007-gf}. The layer m-1, which is also interior, consists
of aragonite and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its
Mg/Ca intensity ratio consistently falls below the range of interest
(0.2 and above). Compared to other patelloid shells, some of whose
interior is almost entirely made of aragonite, this layer is very thin.

@@

\caption{\label{fig-Pat_LIBS}Example LIBS maps and line scans of
analysed specimens. (\emph{a+b}) ORK-LT and (\emph{c}+\emph{d})
QG1-7188-1. Note that the ranges of Mg/Ca intensity ratios are not
identical but are chosen arbitrarily to increase contrast. Areas in grey
consist of low-Mg aragonite and fall outside the chosen colour-range.}

>>>>

\begin{abstract}
This study provides a reassessment of Magnesium to Calcium (Mg/Ca)
ratios in Atlantic limpet shells to determine past sea surface
temperatures (SST). While \emph{Patella depressa} along the Spanish
shoreline has repeatedly produced reliable correlations between SST and
Mg/Ca ratios, this relationship is not the case for other patelloid
species. \emph{Patella vulgata}, \emph{Nacella deaureata}, and \emph{N.
magellanica} have been studied using Mg/Ca ratios with mixed or contrary
results. In this study, we present elemental maps of these three species
together with oxygen isotope ratios (\$\delta\&\textsuperscript{18}O)
that confirm a good relationship with SST. Our dataset also reassesses a
specimen which was previously unsuccessful in providing significant
correlations between \(\delta\)\textsuperscript{18}O values and Mg/Ca
ratios. By reassessing these species and including modern and
archaeological specimens (n=12) from three patelloid species (\emph{P.
vulgata}, \emph{N. deaureata}, and \emph{N. magellanica}) we further add
to the growing set of evidence for the reliable use of Mg/Ca ratios to
detect palaeotemperature change. As a result, these species can in the
future serve to determine ontogenetic age and season of capture as well
as to reveal locations of interest within the growth record (i.e.~annual
temperature minima and maxima) for the targeted analysis using
\(\delta\)\textsuperscript{18}O and clumped isotope analysis.
\end{abstract}

@@

Limpet shells are commonly found within archaeological sites and past
shorelines due to their robust carbonate structure and the long-term use
of limpets as a marine food source \citep{Robson2023-oz}. They have been
successfully studied in the context of coastal subsistence economies and
site occupation on seasonal
\citep{Shackleton1973-ij, Parker2018-wf, Bosch2018-ud} or long-term
scales \citep{Ortiz2015-mr}, as well as archives of palaeotemperature
\citep{Fenger2007-gf, Surge2012-ba, Wang2012-ee, Colonese2012-ct, Ferguson2011-zl}.
Determining past sea surface temperature (SST) change recorded in
patelloid limpet shells largely relies on the measurement of
\(\delta\)\textsuperscript{18}O values within the calcitic parts of the
shell, but attempts have been made to also use elemental ratios, such as
magnesium to calcium (Mg/Ca) ratios
\citep{Graniero2017-io, Ferguson2011-zl}, to have an alternative measure
that potentially provides a more accurate SST estimate than
\(\delta\)\textsuperscript{18}O-values,as the latter proxy is affected
by changes in the mixing of fresh- and saltwater. In addition, the data
acquisition of elemental ratios --- either through
laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser
induced breakdown spectroscopy (LIBS), can be much faster and
cost-effective, increasing the number of specimens that can be studied
overall \citep{Durham2017-fh, Hausmann2023-ih}.

Problematically, links between Mg/Ca ratios and SST changes in many
marine mollusc shells, including limpets, have shown to be unreliable
\citep{Surge2008-ri, Wanamaker2008-zl, Schone2010-yl, Freitas2012-tx, Graniero2017-io, Poulain2015-dg, Vihtakari2017-wd}.
This is particularly the case where there is little available additional
information on metabolic processes, organic components of the shell
matrix, intra-increment and intra-shell variability, and growth rates,
which can independently and unpredictably affect Mg/Ca ratios and
confound their interpretation as a temperature proxy. Confusingly, in
some cases multiple different temperature equations have been found for
the same species (see Freitas \emph{et al}. \citeyearpar{Freitas2012-tx}
and Vihtakari \emph{et al}. \citeyearpar{Vihtakari2017-wd} and
references therein). Coeval specimens sharing one locality can also show
differences in their relation to SST \citep{Hausmann2019-fi}. Where the
use of Mg/Ca ratios as a palaeotemperature proxy was successful,
anomalous patterns in some specimens had to be filtered out by hand,
reducing the overall robustness of the results of those successful
studies \citep{Ferguson2011-zl}.

Recent research particularly of \emph{Patella} spp. in the Mediterranean
and southwest Europe have provided promising results
\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
However, there remains a lack of clarity for Atlantic limpet species,
particularly since it was shown, that the Mg/Ca ratio in their shells
are not reliable recorders of palaeotemperature \citep{Graniero2017-io}.

In this study we will repeat and expand the analysis of Atlantic limpets
to determine the reliability of Mg/Ca ratios as palaeotemperature
proxies. To do this, we sampled a set of previously published and
unpublished limpet specimens dating to modern and archaeological
contexts using LIBS. LIBS allows us to carry out 2D imaging of entire
shell sections, which allows navigation of the complex elemental
structure of the shell and better separate the external,
temperature-related changes from the internal and less understood
factors that influence the Mg/Ca ratio. By working with already
published datasets, we were able to simultaneously avoid costs for new
high-resolution \(\delta\)\textsuperscript{18}O-data, to use real-world
examples, and to also provide pilot-data for areas of existing research
interest. Generally establishing the usefulness and reliability of Mg/Ca
ratios as SST proxy in limpets provides a platform for future research
and an important steppingstone to a better understanding of elemental
ratios in other marine mollusc shells.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2449}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1429}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1633}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2041}}@{}}
\caption{Overview of the modern and archaeological limpet specimens
analysed in this study.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Cambaceres Bay
& ND-1016-3 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & - Tierra del & ND-1016-4 & analysis \\
& & \emph{N. magellanica} & Fuego (AR) & NM-1016-1 & \\
& & & & NM-1016-3 & \\
& & & & & \\
& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - &
ORK-LT5 & \(\delta\)\textsuperscript{18}O and
\(\delta\)\textsuperscript{13}C \\
& & & Westray (UK) & & analysis; \\
& & & & & Mg, Li, Sr, Ca. \\
& & & & & \\
& & & & & \\
Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
Bay - & QG1-7188-1 & Stable oxygen \\
& & & Westray (UK) & QG1-7189-2 & and carbon \\
& & & & QG2-1061-1 & isotope analysis \\
& & & & QG2-1064-1 & \\
& & & & QG2-7180-1 & \\
& & & & QG2-7180-2 & \\
& & & & QG1-7246-1 & \\
\end{longtable}

@@

Specimens were sourced from two main locations: the Beagle Channel in
Tierra del Fuego (Argentina) and the island of West Ray in the Orkney
archipelago of Scotland (UK). Four modern specimens were collected from
the Beagle Channel in Cambaceres Bay in October 2016. The area receives
around 570 mm of precipitation, but the channel waters are predominantly
influenced by marine currents and in the period of October 2015 to
October 2016 stayed around 30.7 ± 0.7 psu (practical salinity units)
\citep{Nicastro2020-ih}. SSTs range from 4.7ºC in August to 10ºC in
February (Figure~\ref{fig-SSTs} -- red line).

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
August 2009 and has been previously analysed for
\(\delta\)\textsuperscript{18}O and \(\delta\)\textsuperscript{13}C
values, as well as LA-ICP-MS \citep{Graniero2017-io}. Westray lies about
70 km north of the Scottish mainland and experiences virtually no
freshwater input with salinity values of 34.0 -- 34.5 psu
\citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in March to
13.8ºC in August (Figure~\ref{fig-SSTs} -- blue line).

@@

We also studied 7 archaeological specimens of \emph{P. vulgata} from the
Quoygrew site on Westray dating to 900 -- 1200 CE. These shells have
previously been studied to characterise seasonal temperature change
during the Medieval Climate Anomaly (810 -- 1229 CE)
\citep{Surge2012-ba}. Together they show a range of lifespans, with some
comparatively long lived for their size (QG1-7246-1, 12 years within 16
mm of growth) making them particularly susceptible to within-shell
time-averaging because of slow growth rates. That said, the majority are
under 5 years old yielding sufficiently fast growth to nearly capture
seasonal amplitudes of SST.

@@

Mg/Ca elemental imaging was carried out using LIBS at the Leibniz
Zentrum für Archäologie (Mainz-Germany) following previously published
methods \citep{Hausmann2023-ih}. The imaging involves the ablation of
the shell section using an infrared (1064 nm) laser (1--2 mJ, 100 Hz)
onto a surface area of 20--30 µm to generate a plasma plume. This plume
emits light, which is measured using a synchronised spectrometer (delay
time = 0.5 µs, integration time = 1 µs). The resulting light spectrum
quantifies the emission lines of Magnesium (MgII; 279.553 nm) and
Calcium (CaII; 315.887 nm) to determine their intensity ratio. While
these two peaks alone do not represent the molar concentrations of both
elements (as opposed to calibration free LIBS; e.g.
\citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
correlated to the molar concentration \citep{Hausmann2017-oa} and can be
used as a reliable indicator of Mg/Ca variation within the shell
carbonate. All values reported here are in arbitrary units based on the
ratio of intensities of the peaks above.

@@

\subsubsection{Oxygen isotope ratios}\label{oxygen-isotope-ratios}

High-quality \(\delta\)\textsuperscript{18}O values were acquired from
previous publications on \emph{P. vulgata}
\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
the stable isotope analysis methodology can be found. In short, all
shells were sectioned along the main growth axis to expose the internal
growth structure. From the exposed section carbonate powder samples were
acquired using a micromill with resolutions of 100--200 µm and by
targeting the calcitic M+2 layer of the shells. Analytical precision of
the isotope values was consistently at 0.05--0.10‰ and values are
reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee
Belemnite) standard.

@@

The elemental imaging of \emph{P. vulgata} shells consistently showed
repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca
intensity ratios indicating and low ratios indicating low temperature
periods. As examples, Figure~\ref{fig-Pat_LIBS} shows the
2D-distribution of elemental ratios across the entire section of ORK-LT5
and the preserved anterior side of QG1-7188-1 (additional maps can be
found in the supplementaries). The repeating patterns are found across
the calcitic layers that are exterior to the myostracum (layers m+2 and
m+3) as well as the layers that are interior (m-2)
\citep{Fenger2007-gf}. The layer m-1, which is also interior, consists
of aragonite and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its
Mg/Ca intensity ratio consistently falls below the range of interest
(0.2 and above). Compared to other patelloid shells, some of whose
interior is almost entirely made of aragonite, this layer is very thin.
\newline

@@

\caption{\label{fig-Pat_LIBS}Example LIBS maps and line scans of
analysed specimens. (\emph{a+c}) ORK-LT and (\emph{b}+\emph{d})
QG1-7188-1. Note that the ranges of Mg/Ca intensity ratios are not
identical but are chosen arbitrarily to increase contrast. Areas in grey
consist of low-Mg aragonite and fall outside the chosen colour-range.}
